---
title: "Group_A8"
author: "Ayesha Gamage-(ayega981)//"
date: "2023-11-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Assignment 1. Handwritten digit recognition with Knearest neighbors.

## Assignment 2. Linear regression and ridge regression.
```{r get library and create data frame,echo=FALSE,warning=FALSE}
      library(caret)
      setwd("D:/MSC/SEMESTER1/732A99 Maching lerning/Exercixe")
      all_data = read.csv("parkinsons.csv")
```

```{r devided data to test and train,echo=FALSE}
        #partition data to train and test (60/40)
        set.seed(12345)
        n = dim(all_data)[1]
        id = sample(1:n, floor(n*0.6))
        train_data = all_data[id,]
        test_data = all_data[-id,]
        
        #data scalling
        datapre = preProcess(train_data, method=c("center", "scale"))
        train_dataS = predict(datapre, train_data)
        test_dataS = predict(datapre,test_data)
        
#linear regression model from the training data 
        lm_model = lm(motor_UPDRS ~ -1+Jitter.Abs.+Jitter.RAP+Jitter.PPQ5+Jitter.DDP+Shimmer+Shimmer.dB.+Shimmer.APQ3+Shimmer.APQ5+Shimmer.APQ11+Shimmer.DDA+NHR+HNR+RPDE+DFA+PPE,train_dataS)
        summary(lm_model)
        para_vector = as.vector(coef(lm_model))
        
 # predict test data and training data, fit training data to model
        train_predictions <- predict(lm_model, newdata = train_dataS)
        test_predictions <- predict(lm_model, newdata = test_dataS)
        
        
#find MSE training data and test data
        tran_MSE <- mean((train_dataS$motor_UPDRS - train_predictions)^2)
        test_MSE <- mean((test_dataS$motor_UPDRS - test_predictions)^2)
        cat("Training MSE :",tran_MSE)
        cat("Test MSE :",test_MSE)
```

Significantly contributed variables found by examine the summery of lm model. In the summery Significant codes given which variable significantly contribute to model. As well as we can get idea about significantly contributed variables by examine the  coefficients and their associated p-values in the summary of model. The p-value indicates the statistical significance of each variable and low p-value (less than 0.05) are considered statistically contributed.

In this example following variables are significantly contributed to prediction,
Jitter.Abs. ,Shimmer,Shimmer.APQ5,Shimmer.APQ11 ,NHR ,HNR ,DFA,PPE

```{r 3,echo=FALSE}
loglikelihood  = function(teta,sigma,data){
        formula = motor_UPDRS ~-1+ Jitter.Abs.+Jitter.RAP+Jitter.PPQ5+Jitter.DDP+Shimmer+Shimmer.dB.+Shimmer.APQ3+Shimmer.APQ5+Shimmer.APQ11+Shimmer.DDA+NHR+HNR+RPDE+DFA+PPE
        x_matrix = model.matrix(formula,data)
        #x_matrix = as.matrix(data[,7:21])
        t<-(as.matrix(teta))
        return ((-(n/2)*log(2*pi*sigma^2))-((1/2*sigma^2)*sum(((x_matrix %*% t) -data$motor_UPDRS)^2)))
        
Ridge <- function(par, data, lambda) {
       theta = par[1:length(par)-1]
       sigma = par[length(par)]
      #get negative log-likelihood from likelyhood function
       neg_likelihood <- -loglikelihood(theta, sigma, data)
       ridge_penalty <- lambda * sum(theta^2)
       total_neg_likelihood <- neg_likelihood + ridge_penalty
       return(total_neg_likelihood)
}

RidgeOpt = function(lambda){
      x_matrix = as.matrix(train_dataS[,7:21])
      init = c(rep(0, ncol(x_matrix)),0.1)
      result = optim(par = init,fn = Ridge, data = train_dataS, lambda = lambda,method = "BFGS")
      return(result)
}

 find_MSE = function(parameters ,datafm){
       parama = as.matrix(parameters$par[1:length(parameters$par)-1])
       predct_train = as.matrix(datafm[,7:21])
       pred = predct_train%*% parama
       return(mean((datafm$motor_UPDRS - pred)^2))
 }

 
 DF = function(lambda) {
      opt_result = RidgeOpt(lambda)
      theta =  as.matrix(opt_result$par[1:length(opt_result$par)-1])
      X = as.matrix(train_dataS[,7:21])
      H <- X %*% solve(t(X) %*% X + lambda * diag(length(theta))) %*% t(X)
      df <- sum(diag(H))
      return(df)
 }
  
 train_MSE_1 = find_MSE(opt_result1,train_dataS)
 train_MSE_100 =find_MSE(opt_result100,train_dataS)
 train_MSE_1000 =find_MSE(opt_result1000,train_dataS)

 test_MSE_1 = find_MSE(opt_result1,test_dataS)
 test_MSE_100 =find_MSE(opt_result100,test_dataS)
 test_MSE_1000 =find_MSE(opt_result1000,test_dataS)

 cat("\u03BB\ :1     training_MSE:",train_MSE_1, "\n")
 cat("\u03BB\ :100   training_MSE:",train_MSE_100, "\n")
 cat("\u03BB\ :1000  training_MSE:",train_MSE_1000, "\n")
 cat("\u03BB\ :1     test_MSE:",test_MSE_1, "\n")
 cat("\u03BB\ :100   test_MSE:",test_MSE_100, "\n")
 cat("\u03BB\ :1000  test_MSE:",test_MSE_1000, "\n")
}
```

